Transposons are a major contributor to gene expression variability under selection in rice populations

Version 1

Abstract

Abstract

Transposable elements are an important source of genome variability. Here, we analyze their contribution to gene expression variability in crops by performing a TE insertion polymorphism (TIP)-eQTL mapping using expression data from rice varieties from the O. sativa ssp. indica and O. sativa ssp. japonica subspecies. Our data shows that TE insertions are associated with changes of expression of many genes known to be targets of rice domestication and breeding. An important fraction of these insertions were already present in the rice wild ancestors, and have been differentially selected in indica and japonica rice populations. Taken together, these data show that small changes of expression in signal transduction genes induced by TE insertions accompany the domestication and adaptation of rice populations.

Introduction

Transposable elements (TEs) are a major component of eukaryotic genomes, particularly in plants 1. Their ability to move, creating mutations by insertion/excision, and amplify in genomes - thus generating new copies that provide opportunities for recombination - make them a major source of genetic variability 2. For this reason TEs are considered a major driver of plant genome evolution, both in the wild or under human selection 3. TE insertions and other structural variants, however, have been frequently overlooked when using genome wide association studies (GWAS) to look for genetic variants linked to interesting phenotypes 4. The recent development of efficient tools to identify transposon insertion polymorphisms (TIPs) 5 has allowed the incorporation of TIPs in these analyses, and several GWAS reports have shown that TIPs uncover additional variability linked to phenotypic traits 6–8. Interestingly, TIPs often explain more phenotypic variance compared to SNPs, and can be used for genomic prediction 9. The reason for this could be that, as compared with SNPs, TE insertions are more frequently the causal mutation.

A major fraction of the mutations linked to crop domestication and breeding are associated with changes of the expression of genes involved in signal transduction 10,11. TEs can alter gene transcription, activating or repressing them, by different means. On the one hand, insertion of TEs within a promoter can interfere with transcription, and the silencing of TEs inserted close to genes can result in repression of gene expression 12–14. Alternatively, TE insertions can also result in gene overexpression 15,16, as TEs bring their own transcriptional promoters that can induce expression of neighboring genes. This is particularly clear for LTR-retrotransposons which contain promoters in their 5′ and 3′ LTRs, and may provide nearby genes with alternative promoters. In addition, TEs can also contain transcription factor binding sites (TFBS) that can alter the expression of host genes 17; it has been shown, for example, that MITEs have frequently amplified and redistributed TFBS in plant genomes 18.

Here we explore the impact of the movement of TEs during the recent evolution of rice on the variability of gene expression in this species. We took advantage of a recently published transcriptional analysis of 208 rice varieties grown under wet paddy and intermittent drought conditions 19 to perform an expression quantitative trait locus (eQTL) GWAS using TIPs as genetic markers. We show that TIPs are frequently associated with changes of expression of rice genes and that many TIPs altering the expression of regulatory genes have been positively selected in indica and japonica rice subspecies.

Characterization of rice TIP-eQTLs.

A-B, Association between TIPs and gene expression levels in cis (TIP-eQTLs) found in indica and japonica population along the twelve rice chromosomes. Horizontal lines represent 5% significance thresholds corrected for multiple testing using FDR method. C, Number of TIP-eQTLs found in each condition classified at the order TE level. White bars represent the number of total TIPs (secondary axis). D, Expression variance explained by leading TIP-eQTL and SNP-eQTL. Each point represents a gene. Indica and Japonica TIP-eQTLs are combined into a single plot. Number of TIP-eQTLs (E) or SNP-eQTL (F) per Mb of gene feature. Upstream and downstream regions are 1 Kb long. Gene region includes 5’ and 3’ UTRs. All the TIPs and SNPs with a significant association (FDR < 5%) were used in this analysis.

Results

TIPs are associated with gene expression variation in rice

We used a recently published dataset containing genome resequencing and transcription data for rice (Oryza sativa) varieties to perform TIP-eQTL mapping analyses. Rice is one of the most important food crops in the world, and varieties are generally found in two subspecies – O. sativa ssp. japonica and ssp indica. Other minor variety groups include aus varieties, which appear closely related to indica, and basmati rice varieties that appear to be a hybrid between ssp. japonica and aus. The data we analyzed consists of an ‘indica’ dataset for 126 O. sativa ssp. indica and some circum-aus accessions, and 84 ‘japonica’ accessions comprising of O. sativa ssp. japonica and some circum-basmati varieties, described in Groen et al. (2020) 19. We predicted TIPs using Popoolation TE2 20, and identified 45,050 TIPs. Using the Minghui MH63 short read data and assembled draft reference genome 21 we estimated the performance of the TIP genotyping on this dataset to have 97.1% sensitivity and 92.2% precision (Supplementary text 1). Most of the TIPs detected are related to DNA transposons (84%), especially MITEs belonging to the Stowaway (24%) and Tourist (20%) superfamilies; LTR retrotransposons make up a minor fraction (9 %).

The 45,050 TIPs are distributed along the 12 chromosomes and 81.3 % of them are less than 5 kb away from an annotated gene (IRGSP RAP-DB gene models). This results in up to 81.4% of rice genes having a TIP less than 5 Kb distant. In order to analyze the potential of TIPs to generate gene expression variability in rice, we used the 3’ mRNA-seq expression data from leaves of adult plants grown in normally-watered soil (“wet” condition) and under intermittent drought-stress (“drought stress” condition) 19. We selected data from 15,549 genes showing expression in more than 10% of the samples. Due to the strong population structure present in rice we separately analyzed data in the indica (126 varieties) and japonica (82 varieties) datasets. We found a total of 563 significant associations between TIPs and gene expression levels in the indica population and 356 in the japonica population (TIP-eQTLs with MAF > 3%, FDR adjusted p value < 0.05 in at least one replicate) (Figure 1A and 1B, Supplementary Table 2). These involve 477 and 317 genes, in indica and japonica respectively, representing 3.1% and 2% of the expressed genes. The TIP-eQTLs were related to the different TE types (orders 22), and the proportion of TE orders found in the significant TIPs did not differ significantly from their representation across the genome (Figure 1C).

In order to estimate the relative contribution of TIPs and SNPs to gene expression changes, we also performed eQTL mapping analyses using 1 million of randomly chosen SNPs as markers 19. We obtained 4,913 SNP-eQTLs in indica and 3,308 SNP-eQTLs in japonica associated with SNPs located less than 5 kb from the gene; each SNP-eQTL represents the leading SNP-gene association. This corresponds to ~31.6% of the genes in indica and ~21.3 % in japonica. For each gene with a significant association, we compared the proportion of variance (R2) explained by the most significantly associated TIPs and/or SNPs. For most genes that have an associated TIP, either there is no associated SNP or the TIP explains more expression variance than the associated SNP (~62% of the indica TIP-eQTLs and ~73% of the japonica TIP-eQTLs) (Figure 1D). This demonstrates that TIPs uncover genetic associations with changes of expression that are not observed when using SNPs as molecular markers.

The SNP-eQTLs found in our study are associated to both positive (47%) and negative (53%) effects on gene expression levels, very close to an even distribution. In contrast, TIPs more frequently (59%) have a negative effect. The proportion varies among TE superfamilies, with TIR/MULEs being close to 50%, and LTR/Gypsy having the most frequently negative impact on gene expression (67%, Supplementary Table 1). This suggests that, as compared with SNP-eQTLs, TIP-eQTLs are more frequently the likely causal mutation responsible for the change in expression, and that the effect of the insertion is more generally negative, in line with recent analyses done in Capsella 23, especially for long elements such as LTR-retrotransposons.

An analysis of the TIP and SNP-eQTL location with respect to the associated genes shows that while SNP-eQTLs are more evenly distributed across the genic region (Figure 1F), TIP-eQTLs are over-represented in the 1 kb region upstream of the gene (Figure 1E), which is the region that most frequently contains promoter elements regulating transcription. This suggests that although some TIP-eQTLs could be in linkage disequilibrium with the causal mutation for the change of expression, an important fraction of TIP-eQTLs could be the actual causal mutation inducing expression variation.

It has been proposed that an important fraction of gene expression variation is deleterious, and therefore, alleles associated with major expression effects should be maintained at lower frequencies 24,25. An analysis of the effect size of the different TIP-eQTLs taking into account their frequency in the population shows that, indeed, low and high-frequency TIPs (corresponding to rare alleles) show the highest positive or negative effect on the expression of the associated gene (Figure 2 A-D, Supplementary Figure 1). This is consistent with what is described as the rare variant effect 25 where rare (and likely deleterious) mutations in a population are associated with greater effects on gene expression in a population.

Effect size and population frequencies of wet and drought-stress TIP-eQTLs. representation of the effect size (beta) of the TIP-eQTL with respect to the frequency of the insertion TIP-eQTLs in indica (A, C) and japonica (B,D) for positive (A,B) and negative (C,D) effects.

Venn diagram illustrating the intersection between TIP-eQTLs detected in each condition of the two subspecies analyzed, indica (ind), and japonica (jap) (E). Percentage of subspecies-specific TIP-eQTLs, considering as shared the TIPs that fall in the intersection between all the FDR-corrected TIP-eQTLs found in a given population and all the associations of the other population using a relaxed cutoff (p < 0.05, no FDR correction) (F). Percentage of condition-specific TIP-eQTLs considering as shared the TIPs that fall in the intersection between all the FDR-corrected TIP-eQTLs found in a given population and all the associations of the other population using a relaxed cutoff (p < 0.05, no FDR correction. Relationship between the frequencies in indica and japonica populations of the 33,389 non-eQTL TIPs (H) and 829 TIP-eQTLs (I).

A comparison of TIP-eQTLs associated with changes of expression in the two different growth conditions (wet and drought stress) shows that there is an important overlap of eQTLs between the two environments (38.5 % for indica TIP-eQTLs and 33.2% for japonica TIP-eQTLs) (Figure 2E). Such overlap increases to 90% (average of indica and japonica, wet) and 71 % (average of indica and japonica, drought stress) when we compare TIP-eQTLs of one condition with a non-FDR-corrected TIP-eQTLs in the other environment (Figure 2G). This result suggests that the majority of TIP-eQTLs are associated with changes of expression in both growth conditions, and that only a small number of TIP-eQTLs (81 in indica and 78 in japonica) are associated with stress-specific changes of expression. Nevertheless, among the TIP-eQTLs associated with changes of gene expression in both wet and drought stress, there are TIP associated with genes known to be involved in drought tolerance (Supplementary Table 2). As an example, TIP_49046 is associated with a reduction of the expression of the gene synaptotagmin-5 (OsSYT-5). OsSYT-5 encodes a Ca2+ sensing protein with a C2 domain, is expressed in both stressed and non-stressed plants, and it has been recently shown that its silencing enhances drought tolerance in rice 26. Other examples are TIP_23764 and TIP_52367, associated with an increased expression of the gene encoding the OsSAPK10 ABA-activated protein kinase and the S-type euonymus-related lectin gene OsEULS2 in respectively, two genes known to mediate drought stress in rice 27,28

TIP-eQTLs are present at different population frequencies in indica and japonica

Although the overlap of TIPs associated with changes in expression under wet and drought stress conditions is very high, the overlap between indica and japonica TIP-eQTLs is low, and only 90 TIP-eQTLs out of the 563 from indica and 356 from japonica are significantly associated with variation of gene expression in both subspecies (Figure 2E). Even when we make the less-stringent comparison of the TIP-eQTLs of one subspecies with the non-FDR-corrected TIP-eQTLs of the other, as much as ~69% of indica and ~60% of japonica TIP-eQTLs appear as subspecies-specific (Figure 2F). This result may suggest differences in the transcriptional networks in indica and japonica or, alternatively, that different alleles showing expression level differences may be differentially represented in the two subspecies. An analysis of the frequencies in indica and japonica of all the TIP-eQTLs described here shows that approximately one-third of the TIPs are only found in one of the two subspecies. This suggests that TEs have been actively inserting (or have been excised or eliminated from the population) very recently during rice evolution, likely post-domestication and during the crop diversification phase, as has already been proposed for some rice TE families 29,30. Here we show that some of these recent TIPs are associated with changes in gene expression and may therefore have phenotypic consequences. We found 44 TIP-eQTLs corresponding to TIPs present only in japonica varieties and 64 TIP-eQTL corresponding to TIPs present only in indica varieties. Some of these insertions are found at high frequencies in its corresponding subspecies, which suggest that they may have been under positive selection (Supplementary Table 3).

However, two thirds (66%) of all TIP-eQTLs found in this study correspond to TIPs present in both subspecies, which suggests that these insertions are relatively old and were already present in the ancestor of indica and japonica rice. Remarkably, in most cases they are found at very different frequencies in the two subspecies (Figure 2I). This suggests that these insertions were already present in the ancestor of indica and japonica rice and have been differentially retained in the two subspecies. Interestingly, an analysis of all non-eQTL-TIPs shows that an important fraction is found at the same frequency in both genomes (Figure 2H), suggesting that the preferential retention of TIP-eQTLs in indica or japonica may be a specific phenomenon potentially linked to their impact on the expression of the associated genes.

We examined whether TIP-eQTLs present in both indica and japonica varieties are also found in either Oryza rufipogon and Oryza nivara, the wild rice relatives believed to be the ancestors of domesticated rice 31. We looked for the presence of the TIPs identified in rice in a set of 72 accessions of O. rufigogon and 10 accessions of O. nivara (collectively called rufipogon/nivara from now on) (Supplementary Table 4). Up to 552 of the 829 TIP-eQTLs present in indica and/or japonica (66%), can be found in rufipogon/nivara varieties, confirming that an important fraction of TIP-eQTLs found in indica and japonica may be relatively old insertions already present in the wild ancestors of domesticated rice. Not surprisingly, in most cases the TIP frequencies in rufipogon/nivara are different from the frequencies of these insertions in indica and/or japonica (Supplementary Table 5); these may have arisen either due to selection or genetic drift accompanying the bottleneck associated with crop evolution.

In order to identify possible selection of TIPs in indica and japonica populations, we used the Population Branch Statistic (PBS) method 32 which examines strong differentiation in frequencies between populations. This approach consists in comparing the three pairwise FST values between indica, japonica and rufipogon/nivara to estimate the frequency change in TIPs that occurred since the divergence of the two rice subspecies from its wild ancestor. We calculated PBS values for a total of 11,698 TIPs present in the japonica, indica and rufipogon/nivara populations, which included 354 TIPs that are eQTLs in in indica and/or japonica populations. TIP-eQTLs (Figure 3A) showed much higher PBS values in comparison to those not identified as eQTLs (Figure 3B) (mean absolute PBS of 0.06 vs 0.02, respectively, p < 0.01, Wilcoxon test). Furthermore, TIPs with the most extreme PBS values (above 95 percentile) are enriched for TIP-eQTLs (Fisher’s Exact test odds ratio = 4.87, p < 0.01), suggesting that a high fraction of TIP eQTLs have been positively selected in indica or japonica. Note that some of the TIPs not identified as eQTLs in this particular dataset may actually be linked to variation of gene expression in organs, developmental stages or environmental conditions different from the ones analyzed here. Interestingly, up to 35% of the TIP-eQTLs identified here, associated with 119 genes, fell within the top 10% of absolute PBS values for all TIPs (Fig 3A), suggesting that TIPs associated with gene expression variation are likely under differential selection during rice evolution.

Signatures of positive selection on TIP-eQTLs.

A, Absolute PBS of 354 TIP-eQTLs (left) and 11,344 TIPs (no-eQTL, right) present in indica, japonica and rufipogon populations. Dotted vertical lines represent the 95th an 99th percentile of the PBS values of the whole dataset (11,698 TIPs). Red dots represent two examples of TIP-eQTLs with extreme PBS values. B) Population frequency of the two TIPs with extreme PBS values, marked as red dots in the left panel A, as well as of the whole TIP dataset. C) Fst-based tree of the two TIPs with extreme PBS values, as well as of the whole TIP dataset (average Fst).

We observe that TIPs can show evidence of selection in japonica, indica or both subspecies. As examples, Figure 3B and C shows the representation of the PBS analysis, and the frequency in the populations, of two TIP-eQTL whose frequency has greatly increased in indica (TIP_53500), or japonica (TIP_72732), compared with the mean PBS for all TIPs.

Gene variants selected during rice evolution: some examples

Genes linked with TIP-eQTLs showing extreme indica or japonica PBS metrics are good candidates for genes underlying adaptation between different rice subspecies. Among the genes with TIP-eQTLs that have extreme PBS values in indica or japonica we can find several examples that are known to regulate plant architecture, plant and grain development or abiotic stress responses.

Some of the most extreme PBS values are for TIPs associated with changes of expression of genes involved in the signal transduction of hormones, including brassinosteroids, ABA, ethylene, jasmonic acid (JA) and auxin (Supplementary Table 5). We found four different TIP-eQTLs with high PBS values associated with the EG2/OsJAZ1 gene (Os04g0653000), a locus encoding a JA signaling repressor 33; the four insertions are related to MITEs of the Tourist transposon superfamily. Three of the insertions (TIP_32891, TIP_32892 and TIP_32894), located upstream (3.4 kb and 2.7 kb) and downstream (150 bp) of the gene, are associated with an increase of expression of EG2, and are physically linked to one another (mean r2 value = 0.88). In contrast, the fourth insertion (TIP_32893), is not linked to them, is within the first intron of the gene and is associated with expression reduction.

The recent advances in characterizing the pangenome of rice and the super-pangenome that includes its wild ancestors 34–36 has allowed us to characterize the locus in indica, japonica and rufipogon/nivara. This analysis showed that EG2 is primarily found in two different haplotypes (Figure 4A). One haplotype, Hap A, contains the three MITE insertions associated with higher expression of EG2 (Figure 4A, 4B, and Supplementary Figure 2A), while the second haplotype, Hap B, contains the fourth MITE insertion and is associated with reduced EG2 expression (Figure 4A, 4B and Supplementary Figure 2A). Hap A was identified to be at high frequency in rufipogon/nivara (48%), whereas Hap B is at lower frequency (16%). The frequency of both Hap A and Hap B is slightly increased in japonica (65% and 28%, respectively). In indica, however, there is a clear reduction in the frequency of Hap A and a concomitant strong frequency increase of Hap B, associated with reduced expression in indica (16% and 83%, respectively). Interestingly, EG2/OsJAZ1 is a repressor of spikelet development 33 and the number of differentiated spikelets per panicle tends to be lower in japonica as compared with indica 37.

Selection on TIP-eQTLs associated with EG2 expression.

A, Representation of the two main EG2 haplotypes present in rice and rufipogon/nivara populations identified in the rice super-pangenome. Conserved nucleotide regions are connected by gray marks. Structural variants longer than 50bp are shown as white spaces. TIP-eQTLs are shown as red boxes. Additional TIPs are shown as white boxes. B, Boxplot representation of the expression of the two EG2 haplotypes in the indica population. Numbers inside boxplots represent the number of accessions in each group. Analysis of nucleotide diversity (π) and diversity in haplotypes homozygosity tracts (H12) for Hap A (C, E) and Hap B (D, F) in japonica (C, D) and indica (E, F) populations. The vertical dotted black line shows the position of the TIP insertion. the EG2 gene is schematically shown in blue. The horizontal dashed black line represents the mean of 1000 random permutation pulls (for p-value see Supplemental Table 6).

We looked more closely for signs of selection at this specific locus, by examining levels of nucleotide diversity (π38) and diversity in haplotypes homozygosity tracts (H12 39). Consistent with the expression (and possible spikelet phenotypic differences), we found that in japonica the haplotype associated with increased EG2 expression (Hap A) was under positive selection, whereas the haplotype associated with decreased expression of EG2 (Hap B) did not show any sign of selection; this indicates that positive selection acts on individuals carrying the haplotype for high expression of EG2 in japonica (Figure 4C-D). Specifically, we observe significantly lower π and higher H12 in Hap A, associated with increased EG2 expression, while this pattern was not significant for Hap B (supplementary Table 6). However, in indica the opposite pattern appears to prevail, wherein there is a stronger evidence of selection for Hap B, but less so for Hap A (Figure 4E-F). This suggests positive selection towards higher expression of EG2 in japonica, and lower expression of EG2 in indica, which could potentially explain the higher number of spikelets per panicle observed in the latter37.

Another good candidate for genes underlying adaptation of rice is the OsGAP gene. There are two different TIP-eQTLs with high PBS values associated with changes of expression of the OsGAP gene (Os07g0500300) in japonica, whereas they do not correlate with changes in expression in indica (Supplementary Figure 2B). OsGAP encodes a putative GTPase activating protein, similar to CAR proteins (C2-domain abscisic acid-related proteins), that play an important role in ABA signal transduction in Arabidopsis 40. The first TIP (TIP_50057) is located ~4 kb upstream of the OsGAP gene and is associated with an increase of expression of the gene in japonica whereas the second insertion (TIP_50059) is located within the first intron and is associated with a decrease in expression in japonica (Figure 5A-B). The analysis of the locus using the super-pangenome data shows that OsGAP is found in two different haplotypes, one containing TIP_50057 (Hap A) and the other containing TIP_50059, together with some additional structural differences (Hap B; Figure 5A). The two TIPs defining the two haplotypes are present in the rufipogon/nivara population at complementary frequencies (73% and 19%, respectively). Interestingly, the proportion of the two haplotypes in cultivated rice seem to be very different, with an increased frequency of the haplotype associated with a reduced expression of OsGAP (Hap B) in both japonica (~50%) and indica (~95%) were it reaches near fixation.

Selection on TIP-eQTLs associated with OsGAP expression.

A, Representation of the two main OsGAP haplotypes present in rice and rufipogon/nivara populations identified in the rice super-pangenome. Conserved nucleotide regions are connected by gray marks. Structural variants longer than 50bp are shown as white spaces. TIP-eQTLs are shown as red boxes. Additional TIPs are shown as white boxes. B, Boxplot representation of the expression of the two OsGAP haplotypes in the japonica population. Numbers inside boxplots represent the number of accessions in each group. Analysis of nucleotide diversity (π) and diversity in haplotypes homozygosity tracts (H12) for Hap A (C) and Hap B (D) in japonica population. The vertical dotted black line shows the position of the TIP insertion. The OsGAP gene is schematically shown in blue. The horizontal dashed black line represents the mean of 1000 random permutation pulls (for p-value see Supplemental Table 6).

It has been proposed that OsGAP is a negative regulator of ABA signaling in seed germination and dormancy, and that reduced expression of OsGAP may prevent rice pre-harvest sprouting (PHS) 40. Reduced seed dormancy is a common target for selection of cultivated varieties, but this trait may come at a cost of a higher PHS risk, which may be a problem, especially in regions where heavy rain is common during the harvest season. Therefore, the appropriate degree of dormancy may depend on the agroecological conditions in which a particular variety is grown. We see here that two independent TIPs are associated with changes of OsGAP expression in japonica but not in indica (Figure 5A, Supplementary Figure 2B). Interestingly, there are clear and strong signs of selection for increased expression of OsGAP in japonica, wherein Hap A was under positive selection, and Hap B was selectively neutral, indicating positive selection acts on only those individuals carrying haplotype for high expression of OsGAP in japonica (Figure 5C,D; Supplementary Table 6). In contrast, there is no evidence for selection in indica varieties despite the near fixation of one haplotype. This suggests that an increased level of OsGAP may be relevant for the control of dormancy in japonica but not in indica (despite its near-fixation in the latter), which would be in line with recent data suggesting that seed dormancy is regulated by different genes/alleles in indica and japonica 41.

Finally, we find a TIP-eQTL whose frequency has greatly increased in cultivated rice with respect to rufipogon/nivara. TIP_45706, corresponding to an insertion of a gypsy-like LTR-RT at ~1 kb upstream of the OsMPH1 gene (Figure 6A). This insertion is associated with a decrease in the expression of OsMPH1 in indica and possibly japonica varieties (Figure 6B, Supplementary Figure 2C), and is found at low frequency in rufipogon/nivara (~11%) but at high frequency in both indica (70%) and japonica (95%). OsMPH1 encodes a MYB-like transcription factor that has been shown to regulate plant height, its reduced expression resulting in shorter plants 42. A reduction in plant height due to a mutation in the SD1 gene, which encodes the gibberellin biosynthesis gene GA-20ox, was at the origin of the Green Revolution in the 1960s, but it has been shown that alleles of SD1 resulting in shorter culm length were also selected during the domestication of japonica rice 43. Our results suggest possible parallel selection of alleles in other genes that may lead to similar phenotypes. This is further reinforced by strong selection acting on the haplotype containing the TIP_45706 insertion in both indica and japonica (Figure 6C and 6D Supplementary table 6), leading to reduced expression of OsMPH1 and thus shorter plants.

Selection on TIP-eQTLs associated with OsMPH1 expression.

Representation of the two main OsMPH1 haplotypes present in rice and rufipogon/nivara populations identified in the rice super-pangenome. Conserved nucleotide regions are connected by gray marks. Structural variants longer than 50bp are shown as white spaces. TIP-eQTLs are shown as red boxes. B, Boxplot representation of the expression of the two OsMPH1 haplotypes in the japonica population. Numbers inside boxplots represent the number of accessions in each group. Analysis of nucleotide diversity (π) and diversity in haplotypes homozygosity tracts (H12) for Hap A in indica population (C) and Hap A (D) in japonica population. The vertical dotted black line shows the position of the TIP insertion. The OsMPH1 gene, which is schematically shown in blue. The horizontal dashed black line represents the mean of 1000 random permutation pulls (for p-value see Supplemental Table 6).

Discussion

Transposons are a major source of genome variability, and are known to affect gene expression in numerous ways. The importance of rare variants and unfixed TE insertions for gene expression variation in humans 44 and plants 23 has already been demonstrated. A recent pangenome analysis in tomato has shown that the phenotypic variation of important crop traits is linked to structural variants present in the species, most of which are actually TIPs, which are associated to subtle changes of transcription of genes involved in signal transduction 45. In the present study we evaluated the importance of TIP-related expression variability in the recent evolution of rice. To this end we performed a TIP-eQTL mapping using expression data from rice varieties from the O. sativa ssp. indica and O. sativa ssp. japonica subspecies. We show here that using TIPs in addition to SNPs as genetic information allows the uncovering of additional genomic associations to changes in gene expression, and that when both TIPs and SNPs are both associated, TIPs often explain more of the variance in expression. This is in line with recent reports that have also used TIPs for GWAS with different phenotypes in tomato and rice 7,8, and suggests that TIPs may be more frequently the causal mutation of a particular phenotype than SNPs. The concentration of TIP-eQTLs in upstream regions of genes, where most transcriptional regulatory elements usually reside, also suggest that a high fraction of the TIP-eQTLs here described are the actual mutation underlying gene expression variation.

The close association of some TEs with genes could also partially explain the frequent association of more than one TIP with changes of expression of particular genes. Indeed, among the 718 genes with TIP-eQTLs described here, 18% have two of more TIPs associated with changes of expression. However, not all genes have the same number of TIPs. Among the 30 genes with TIPs in indica and japonica that have the highest population branch statistics (PBS) – an indicator of strong differential selection - 13 have more than one TIP-eQTL. This suggests that TIPs are a major source of gene expression variability in rice, in particular in genes that may have been strongly selected during the recent evolution of rice. In some cases, the different TIPs linked to a gene are associated with opposite effects on its expression and are present in different haplotypes. In these instances, as shown for the OsJAZ1 and OsGAP genes, the two haplotypes have strong signs of selection, although one positively and the other negatively, pointing to role these TE insertions may play in adaptive evolution of these crop subspecies.

The results presented here show that most TIPs associated with variation of gene expression in indica and/or japonica are relatively old insertions that are also present in rufipogon/nivara, the wild ancestors of rice. This is in line with recent data showing that a high number of the structural variants associated with changes of expression in tomato were already present in its wild ancestor 45 and highlights the importance of the standing variation, already present in the wild ancestors, for crop adaptation. It is assumed that most TE insertions are selectively neutral or slightly deleterious 46, and the presence of TIPs associated with expression variation of genes in the wild ancestors of rice, suggests that this variation can be well tolerated in wild rice. Interestingly, we show here that many of these TIPs have been positive or negative selected in rice populations, which shows that they translate into selectable phenotypic differences in the agroecological conditions of cultivated rice. Indeed, we show examples of selected variants with modified expression of genes known to be linked to important traits that were targets of selection. It has been already proposed that a major fraction of the mutations linked to crop domestication and breeding are associated with changes of gene expression involved in signal transduction 10,11. Here we show that specific expression variants of genes involved in signal transduction have been differentially selected in indica and japonica rice populations. In addition, our results also point to TEs as a major driver of gene expression variation selected during crop adaptation and breeding.

Methods

TIP detection

Re-sequencing data for 126 indica and 82 japonica rice accessions was obtained from Groen et al 19 (Bioproject accessions PRJNA557122, PRJNA422249 and PRJEB6180). BBDuK (https://sourceforge.net/projects/bbmap/) was used for adaptor and quality trimming. Clean reads were aligned to the Nipponbare reference genome 47 using BWA 48. PoPoolationTE2 20 was used to detect TIPs in the mode “joint” using Nipponbare TE annotation described by Ou et al 47. TIPs with a zygosity lower than 0.25 in all samples were excluded to avoid false positives. TIPs were further filtered using the parameters –-min-count 5, --max-otherte-count 2, --max-structvar-count 2, and only those having MAF higher than 3% and no missing data were kept. Finally, the TIP matrix was transformed to binary form using zygosity cutoff of 0.05 to define an insertion as present.

TIP and SNP-eQTL mapping

Transcriptome data for the 208 rice accessions (wet and drought stress conditions; tree independent replicates per condition) was obtained from Groen el al 19 and separated into individual replicate matrices. Transcripts expressed in more than 10% of the samples on each replicate were extracted and counts were normalized using the yst function of DESeq2 49. TIP and SNP-eQTL mapping were performed using Matrix eQTL software 50, applying the simple linear regression model and including subpopulation groups as covariates. We used a cut-off distance of 5,000bp to identify cis-eQTLs and a 5% False Discovery Rate threshold for multiple testing corrections.

Selection analyses

For the Population Branch Statistics (PBS) analysis we looked for the presence of rice TIPs in the 82 accessions belonging to the rufipogon/nivara population (Supplementary Table 4) and retained only those present in the three populations (indica, japonica, rufipogon/nivara), resulting in a matrix of 13,622 TIPs and 288 accessions. The TIP matrix was filtered to remove TIPs with more than 5% missing data, or with MAF < 5%. The remaining missing data (1.7%) was imputed using the “wright” algorithm implemented in SNPready. The clean matrix (11,698) was used to calculate the individual PBS values per TIP, following the formulas described in Yi et al., (2010)32.

Along with PBS, to evaluate whether TIP-eQTLs show signs of selection, we estimated nucleotide diversity (pi 38), which is a SFS (site-frequency spectrum) measure to test for the presence of selection. This was done using a custom script in a window of 25SNPs with a sliding window of 5. Further, we also estimated the H12 homozygosity statistic 39 that can detect the presence of both hard and soft sweeps associated with selection. Since we expect long runs of homozygosity in the regions around the selected loci, H12 was calcuted in windows of 50 SNPs, with a sliding window of 25. Both these statistics were estimated separately for indica and japonica, and for the samples with TIP insertions present and absent. This was for a range of 100kb (50kb up-and downstream the TIP insertion), using SNPs identified from Groen et al 19 but without the 1000bp linkage thinning. To test for the significance of the statistics, we performed a permutation test (N=1000) with random sampling with the same number of individuals that were used to estimate the selection statistics. The statistics were estimated for these 1000 random pulls in the same regions, and using the same window sizes as the original statistics.

eLife Assessment

This valuable study reports on the role of transposable elements in gene expression variation in rice and how TE-associated expression changes could have been selected during domestication. The combination of evidence from linkage studies and selection scans for a subset of insertions is convincing, although it is currently unclear how large the contribution of TEs to gene expression variability compared to other variants is, which should be reflected in the title. The work will be of interest to colleagues working on the role of transposable elements in adaptation and to biologists working on domestication.

Reviewer #1 (Public Review):

For many years it has been understood that transposable elements (TEs) are an important source of natural variation. This is because, in addition to simple knockouts of genes, TEs carry regulatory sequences that can, and sometimes do, affect the expression of genes near the TEs. However, because TEs can be difficult to map to reference genomes, they have generally not been used for trait mapping. Instead, single nucleotide polymorphisms are widely used because they are easy to detect when using short reads. However, improvements in sequencing technology, as well as an increased appreciation of the importance of TEs to both linked to favorable alleles and are more likely to be causing the changes that make those alleles beneficial in a given environment. Further, because TE activity can occur after bottlenecks, they can provide polymorphisms in the absence of variation in point mutations.

In this manuscript, the authors carefully examine insertion polymorphisms in rice and demonstrate linkage to differences in expression. To do this, they used expression quantitative trait locus (eQTL) GWAS using TIPs as genetic markers to examine variation in 208 rice accessions. Because they chose to focus on genes that were expressed in at least 10% of the accessions, presumably because more rare variants would end up lacking statistical power. This is an understandable decision, but it says that recent insertions, such as the MITE elements detailed in a previous paper, would not be included. Importantly, although TIPs associated with differentially expressed genes are far less common than SNPs' traditional eQTLs, there were a significant number of eQTLs that showed linkage to TIPs but not to QTL.

The authors then show that of the eQTLs associated with both TIPs and SNPs, TIPs are more tightly linked to the eQTL, and are more likely to be associated with a reduction in expression, with variation in the effects of various TEs families supporting that hypothesis. Here and throughout, however, the distance of the TEs could be an important variable. It is also worth noting the relative numbers in order to assess the claim in the title of the paper. The total number of eQTL-TIPs is ten-fold less than the number of eQTL-SNPs, and, of the eQTLs that have both, there are a significant number of eQTL-TIPs that are not more tightly linked to the expression differences than the eQTL.

The authors show that eQTL-TIPs are more likely to be in the promoter-proximal region, but this may be due to insertion bias, which is well documented in DNA-type elements. Here and throughout the authors are careful to state that the data is consistent with the hypothesis that TEs are the cause of the change, but do not claim that the data demonstrate that they are.

Throughout the rest of the manuscript, the authors systematically build the case for a causal role for TEs by showing, for instance, that eQTL-TIPs show much stronger evidence for selection, with increased expression being more likely to be selected than decreased expression. The authors provide examples of genes most likely to have been affected by TE insertions.

Overall, the authors build a convincing case for TEs being an important source of regulatory information. I don't have any issues with the analysis, but I am concerned about the sweeping claims made in the title. Once you get rid of eQTLs that could be altered by either SNPs or TIPs and include only those insertions that show strong evidence of selection, the number of genes is reduced to only 30. And even in those cases, the observed linkage is just that, not definitive evidence for the involvement of TEs. Although clearly beyond the scope of this analysis, transgenic constructs with the TEs present or removed, or even segregating families, would have been far more convincing.

The fact that many of the eQTL-TIPs were relatively old is interesting because it suggests that selection in domesticated rice was on pre-existing variation rather than new insertions. This may strengthen the argument because those older insertions are less likely to be purged due to negative effects on gene expression. Given that the sequence of these TEs is likely to have diverged from others in the same family, it would have been interesting to see if selection in favor of a regulatory function had caused these particular insertions to move away from more typical examples of the family.

Reviewer #2 (Public Review):

In this manuscript, Castanera et al. investigated how transposable elements (TEs) altered gene expression in rice and how these changes were selected during the domestication of rice. Using GWAS, the authors found many TE polymorphisms in the proximity of genes to be correlated to distinct gene expression patterns between O. sativa ssp. japonica and O. sativa ssp. indica and between two different growing conditions (wet and drought). Thereby, the authors found some evidence of positive selection on some TE polymorphisms that could have contributed to the evolution of the different rice subspecies. These findings are underlined by some examples, which illustrate how changes in the expression of some specific genes could have been advantageous under different conditions. In this work, the authors manage to show that TEs should not be ignored when investigating the domestication of rise as they could have played an important role in contributing to the genetic diversity that was selected. However, this study stops short of identifying causations as the used method, GWAS, can only identify promising correlations. Nevertheless, this study contributes interesting insights into the role TEs played during the evolution of rice and will be of interest to a broader audience interested in the role TEs played during the evolution of plants in general.

Author Response:

The following is the authors' response to the original reviews.

Reviewer #1 (Public Review):

[…] Overall, the authors build a convincing case for TEs being an important source of regulatory information. I don't have any issues with the analysis, but I am concerned about the sweeping claims made in the title. Once you get rid of eQTLs that could be altered by either SNPs or TIPs and include only those insertions that show strong evidence of selection, the number of genes is reduced to only 30. And even in those cases, the observed linkage is just that, not definitive evidence for the involvement of TEs. Although clearly beyond the scope of this analysis, transgenic constructs with the TEs present or removed, or even segregating families, would have been far more convincing.

We notice that the referee thinks that we "built a convincing case for TEs being an important source of regulatory information". This is what we wanted to convey in the title, were we were cautious to not claiming that TEs are the most important contributor to gene expression variability in rice populations. However, we agree with the referee that the title may be improved to better describe the results presented. We have therefore changed the title to "Transposons are an important contributor to gene expression variability under selection in rice populations".

With respect to demonstrating causality by removing or introducing the TEs, this is indeed a work we plant to do but that, as stated by the referee, is beyond the scope of this analysis.

The fact that many of the eQTL-TIPs were relatively old is interesting because it suggests that selection in domesticated rice was on pre-existing variation rather than new insertions. This may strengthen the argument because those older insertions are less likely to be purged due to negative effects on gene expression. Given that the sequence of these TEs is likely to have diverged from others in the same family, it would have been interesting to see if selection in favor of a regulatory function had caused these particular insertions to move away from more typical examples of the family.

The TIP-eQTL are from different classes, superfamilies and families and the number of TIP-eQTLs of the same family is too small to deduce sequence communalities (4.6 TIP-eQTLs/family in indica and 3.6 TIP-eQTLs/family in japonica). On the other hand the effect of TIPs on expression can be positive or negative (we show actually that it is often negative). In the later case, a plausible scenario would be of the insertion inactivating a promoter element, and in this case it would be the insertion itself, and not the actual sequence of the TE what would be selected.

Also, previous work done in our lab has shown that TEs can amplify and mobilize transcription factor binding sites that are bound by the TF even when they are not close to a gene and therefore probably not directly affecting gene expression (Hénaff et al.,2014. The Plant Journal). In that case, the sequence of the eQTL TEs and those that are far away from genes will not necessarily differ.

Reviewer #2 (Public Review):

In this manuscript, Castanera et al. investigated how transposable elements (TEs) altered gene expression in rice and how these changes were selected during the domestication of rice. Using GWAS, the authors found many TE polymorphisms in the proximity of genes to be correlated to distinct gene expression patterns between O. sativa ssp. japonica and O. sativa ssp. indica and between two different growing conditions (wet and drought). Thereby, the authors found some evidence of positive selection on some TE polymorphisms that could have contributed to the evolution of the different rice subspecies. These findings are underlined by some examples, which illustrate how changes in the expression of some specific genes could have been advantageous under different conditions. In this work, the authors manage to show that TEs should not be ignored when investigating the domestication of rise as they could have played an important role in contributing to the genetic diversity that was selected. However, this study stops short of identifying causations as the used method, GWAS, can only identify promising correlations. Nevertheless, this study contributes interesting insights into the role TEs played during the evolution of rice and will be of interest to a broader audience interested in the role TEs played during the evolution of plants in general.

We agree with the referee that the results presented do not allow concluding on causality, and we have been careful not to pretend they would in the manuscript. We plan to perform analysis of adding or removing TEs by CRIPR/Cas 9 approaches to address this, but, in line with referee's 1 comment, we think this is beyond the scope of this analysis.

----------

Reviewer #1 (Recommendations For The Authors):

Everything that I need to say is provided in the public portion of my review.

Reviewer #2 (Recommendations For The Authors):

Major concerns:

1. The authors compare the proportion of the variance explained by the most significant TIP and SNP on the observed eQLTs associated with TIPs and SNPs. Thereby the authors conclude that TIPs explain more variance than SNPs. If I am not mistaken the GWAS was run separately for TIPs and SNPs, however, I am wondering if running the GWAS on the combined TIP and SNP dataset might be the better way to compare the variance explained by TIPs and SNPs on gene expression differences. It would be nice to see if these results also hold true if a TIP and SNP combined dataset is used as the most significant marker in a GWAS might not be the causal mutation but might just be linked to the causal mutation. Further in the TIP dataset, the number of markers is only 45k and in the SNP dataset, it is 1 000k, which could bias the GWAS toward finding markers that explain more of the variation in the dataset with fewer markers.

We addressed the reviewer concern by using two complementary approaches, whose results are described in the text (lines 119-121) and in the new Figure 1-figure supplement 1.

First, we addressed the concern regarding the independent GWAS for TIPs and SNPs vs a combined strategy. For this, we built new japonica/indica genotype matrices containing all TIP and SNP matrix together and ran eQTL mapping again. Using the same strategy (association + FDR adjust), we found 100% of the previous TIP-eQTLs and 99% of the previous SNP-eQTLs. We repeated the same analysis (proportion of expression variance), and the results were mostly the same (Figure 1-figure supplement 1A).

Second, we addressed the two concerns (combined genotypes and different amount of TIP and SNP markers) using a single approach. SNP matrices were LD pruned using a r2 = 0.9 and later subsampled to the exact number of TIPs (Indica = 30,396, Japonica = 25,168). We verified that these SNPs covered well the 12 rice chromosomes. SNP and TIP genotypes were later merged into a single matrix, and eQTL mapping was repeated for each of the subspecies and conditions using the same parameters as in the previous version of the manuscript. 100 % of the previously reported TIP-eQTL associations were found using this new approach. Nevertheless, we found a very important drop of sensitivity in the SNP-eQTLs (only 15-20% of the previous associations were detected), possibly due to the strong reduction in the number of SNPs (> 95 %), which results in much lower number of markers at < 5Kb from genes). We repeated the analysis of Figure 1D, and observed very similar results (Figure 1-figure supplement 1D). There is a very important number of TIP-eQTL associations that do not coincide with SNP-eQTLs, (74% in indica, 83% in japonica) indicating that TIP-eQTL mapping is complementary to SNP-eQTL mapping as it uncovers additional associations (note that in this case the overlap between TIP-eQTLs and SNP-eQTLs is lower than in the previous analysis due to the lower sensitivity of SNP-eQTL mapping using less markers). In the cases were both a TIP and a SNP coincide as eQTL, TIPs explained slightly more variance than SNPs in both indica and japonica (in 54% of the cases TIP variance > SNP variance).

2. Line 146 to 152: in this section, the authors describe overlaps between TIP-eQTLs in two different growth conditions, however, in the text it is not mentioned if the TIPs have the same effect on gene expression in the two conditions or if the gene expression is up-regulated in one condition but down-regulated in the other. This information would be interesting to have here, especially as the authors go on to say that only a small number of TIP-eQTLs are stress-specific. The same comment also goes for the eQTL overlap described on lines 167 to 170.

We checked the effect type (positive or negative) of TIP-eQTLs in both scenarios (associations shared between wet/dry conditions, and associations shared between subspecies). In both cases, 100 % of the shared TIP-eQTLs have the same effect type in the two conditions or subspecies. We have updated the text accordingly (Lines 55-157 and Lines 179-181)

3. Lines 192 to 196: the authors mention that the frequency of non-eQTL-TIPs was at the same frequency in indica and japonica, which is in contrast to eQTL-TIPs. However, on line 132 it is mentioned that eQTL-TIPs were overrepresented in 1 kb regions upstream of genes. Hence, is the pattern of the frequency of non-eQTL-TIPs being at the same frequency in indica and japonica also observed in the 1 kb regions upstream of genes and/or if the distribution of non-eQTL-TIPs is matched to one of the eQTL-TIPs? Or is this pattern driven by non-eQTL-TIPs far away from genes?

We checked the frequencies of TIPs at 1Kb upstream genes and found that the general pattern is maintained, with the frequencies of TIP no-eQTLs being more correlated than that of TIP-eQTLs. We have included this information (lines 204-206) an added a new supplementary file (Figure 2-figure supplement 2)

4. In the discussion, the authors could briefly discuss how linked selection affecting TIPs could contribute to the observed results. After reading the second example in the result section where one of the example TIPs (TIP_50059) is found on the Hap B which contains "some additional structural differences" (line 290), I was left wondering how much of the increase in TIP frequency can be attributed to genetic hitchhiking? And how much of the results could be caused by linked selection, especially when considering that structural variations are not included in the GWAS analyses.

We agree with the referee in that some of the TIP eQTLs here described might be not the actual cause of expression variability (ej, TIP linked with the causal mutation), although we cannot know the exact fraction. This is stated in several places of the results and discussion sections. However, the fact that TIPs tend to explain more variance than SNPs and that TIP eQTL, but not SNP eQTL, tend to concentrate in the upstream proximal region of genes where most transcription regulatory sequences are located (Figure 1), suggest that TIP eQTLs could be more frequently the causal than SNP eQTLs. We revised the text to ensure that we convey this message appropriately.

Minor comments:

• Lines 80 to 83: the description of the rice phylogeny should be moved to the introduction.

Done (Lines 68-72)

• Line 177 to 186: It was unclear to me if the authors checked in the ancestral rice population laced the TIPs described in this section as recently inserted in the indica and japonica ssp. It would be nice to add this information to this section.

Thanks to the referee comment we noted an imprecision in the text. The approximate 1/3 of subspecies specific TIP-eQTLs refers to the TIPs at 3% MAF (ie, some of these insertions could be present at > 3% in indica, but at < 3% MAF in japonica). We now indicate only the TIPs that are truly specific to any of the two subspecies (frequency is zero in one of the two) and looked for their presence in rufipogon:

59 insertions are indica-specific. Of those, 33 are present in rufipogon.

21 insertions are japonica-specific. Of those, 5 are present in rufipogon.

We have incorporated this information in the manuscript (Lines 185-189). The species-specific TIPs are also available in the Supplementary File 3.

• Line 353: "have two of more TIPs" should be "two or more"

Done (Line 369)

• Figure 1D: Using a square layout instead of a rectangle layout for the plot will make it easier to interpret.

Done.

Version 2

Abstract

Abstract

Transposable elements are an important source of genome variability. Here, we analyze their contribution to gene expression variability in rice by performing a TE insertion polymorphism (TIP)-eQTL mapping using expression data from 208 varieties from the O. sativa ssp. indica and O. sativa ssp. japonica subspecies. Our data shows that TE insertions are associated with changes of expression of many genes known to be targets of rice domestication and breeding. An important fraction of these insertions were already present in the rice wild ancestors, and have been differentially selected in indica and japonica rice populations. Taken together, our results show that small changes of expression in signal transduction genes induced by TE insertions accompany the domestication and adaptation of rice populations.

Introduction

Transposable elements (TEs) are a major component of eukaryotic genomes, particularly in plants [1]. Their ability to move, creating mutations by insertion/excision, and amplify in genomes - thus generating new copies that provide opportunities for recombination - make them a major source of genetic variability [2]. For this reason TEs are considered a major driver of plant genome evolution, both in the wild or under human selection [3]. TE insertions and other structural variants, however, have been frequently overlooked when using genome wide association studies (GWAS) to look for genetic variants linked to interesting phenotypes [4]. The recent development of efficient tools to identify transposon insertion polymorphisms (TIPs) [5] has allowed the incorporation of TIPs in these analyses, and several GWAS reports have shown that TIPs uncover additional variability linked to phenotypic traits [6–8]. Interestingly, TIPs often explain more phenotypic variance compared to SNPs, and can be used for genomic prediction [9]. The reason for this could be that, as compared with SNPs, TE insertions are more frequently the causal mutation.

A major fraction of the mutations linked to crop domestication and breeding are associated with changes of the expression of genes involved in signal transduction [10,11]. TEs can alter gene transcription, activating or repressing them, by different means. On the one hand, insertion of TEs within a promoter can interfere with transcription, and the silencing of TEs inserted close to genes can result in repression of gene expression [12–14]. Alternatively, TE insertions can also result in gene overexpression [15,16], as TEs bring their own transcriptional promoters that can induce expression of neighboring genes. This is particularly clear for LTR-retrotransposons which contain promoters in their 5′ and 3′ LTRs, and may provide nearby genes with alternative promoters. In addition, TEs can also contain transcription factor binding sites (TFBS) that can alter the expression of host genes [17]; it has been shown, for example, that MITEs have frequently amplified and redistributed TFBS in plant genomes [18].

Here we explore the impact of the movement of TEs during the recent evolution of rice on the variability of gene expression in this species. Rice is one of the most important food crops in the world, and varieties are generally found in two subspecies – O. sativa ssp. japonica and ssp indica. Other minor variety groups include aus varieties, which appear closely related to indica, and basmati rice varieties that appear to be a hybrid between ssp. japonica and aus. We took advantage of a recently published transcriptional analysis of 208 rice varieties belonging to the major rice variety groups grown under wet paddy and intermittent drought conditions [19] to perform an expression quantitative trait locus (eQTL) GWAS using TIPs as genetic markers. We show that TIPs are frequently associated with changes of expression of rice genes and that many TIPs altering the expression of regulatory genes have been positively selected in indica and japonica rice subspecies.

Results

TIPs are associated with gene expression variation in rice

We used a recently published dataset containing genome resequencing and transcription data for rice (Oryza sativa) varieties to perform TIP-eQTL mapping analyses. The data we analyzed consists of an ‘indica’ dataset for 126 O. sativa ssp. indica and some circum-aus accessions, and 82 ‘japonica’ accessions comprising of O. sativa ssp. japonica and some circum-basmati varieties, described in Groen et al. (2020) [19]. We predicted TIPs using PopoolationTE2 [20], and identified 45,050 TIPs. Using the Minghui MH63 short read data and assembled genome [21] we estimated the performance of the TIP genotyping on this dataset to have 97.1% sensitivity and 92.2% precision (see methods). Most of the TIPs detected are DNA transposons (84%), especially MITEs belonging to the Stowaway (24%) and Tourist (20%) superfamilies; LTR retrotransposons make up a minor fraction (9 %).

The 45,050 TIPs are distributed along the 12 chromosomes and 81.3 % of them are less than 5 kb away from an annotated gene (IRGSP RAP-DB gene models). This results in up to 81.4% of rice genes having a TIP less than 5 Kb distant. In order to analyze the potential of TIPs to generate gene expression variability in rice, we used the 3’ mRNA-seq expression data from leaves of adult plants grown in normally-watered soil (“wet” condition) and under intermittent drought-stress (“drought stress” condition) [19]. We selected data from 15,549 genes showing expression in more than 10% of the samples. Due to the strong population structure present in rice we separately analyzed data in the indica (126 varieties) and japonica (82 varieties) datasets. We found a total of 563 significant associations between TIPs and gene expression levels in the indica population and 356 in the japonica population (TIP-eQTLs with MAF > 3%, FDR adjusted p value < 0.05 in at least one replicate) (Figure 1A and 1B, Supplementary file 1). These involve 477 and 317 genes, in indica and japonica respectively, representing 3.1% and 2% of the expressed genes. The TIP-eQTLs were related to the different TE types (orders [22]), and the proportion of TE orders found in the significant TIPs did not differ significantly from their representation across the genome (Figure 1C).

Characterization of rice TIP-eQTLs.

A-B, Association between TIPs and gene expression levels in cis (TIP-eQTLs) found in indica and japonica population along the twelve rice chromosomes. Horizontal lines represent 5% significance thresholds corrected for multiple testing using FDR method. C, Number of TIP-eQTLs found in each condition classified at the order TE level. White bars represent the number of total TIPs (secondary axis). D, Expression variance explained by leading TIP-eQTL and SNP-eQTL. Each point represents a gene. Indica and Japonica TIP-eQTLs are combined into a single plot. Number of TIP-eQTLs (E) or SNP-eQTL (F) per Mb of gene feature. Upstream and downstream regions are 1 Kb long. Gene region includes 5’ and 3’ UTRs. All the TIPs and SNPs with a significant association (FDR < 5%) were used in this analysis.

In order to estimate the relative contribution of TIPs and SNPs to gene expression changes, we also performed eQTL mapping analyses using a high density genotype matrix consisting of 824,311 SNPs for Indica and 706,542 SNPs for Japonica [19]. We obtained 4,913 SNP-eQTLs in indica and 3,308 SNP-eQTLs in japonica associated with SNPs located less than 5 kb from the gene (were each SNP-eQTL represents the leading SNP-gene association). This corresponds to ~31.6% of the genes in indica and ~21.3 % in japonica. For each gene with a significant TIP-eQTL association, we compared the proportion of variance (R2) explained by the most significantly associated TIPs and/or SNPs. For most genes that have an associated TIP, either there is no associated SNP or the TIP explains more expression variance than the associated SNP (~62% of the indica TIP-eQTLs and ~73% of the japonica TIP-eQTLs) (Figure 1D). This result could be reproduced by running the association using SNPs and TIPs together (Figure 1-figure supplement 1A), and also by subsampling SNPs to the same number of TIPs (Figure 1-figure supplement 1B). This demonstrates that TIPs uncover genetic associations with changes of expression that are not observed when using SNPs as molecular markers.

The SNP-eQTLs found in our study are associated to both positive (47%) and negative (53%) effects on gene expression levels, very close to an even distribution. In contrast, TIPs more frequently have a negative effect (59%). The proportion varies among TE superfamilies, with TIR/MULEs being close to 50%, and LTR/Gypsy having the most frequently negative impact on gene expression (67%, Supplementary file 2). This suggests that, as compared with SNP-eQTLs, TIP-eQTLs are more frequently the likely causal mutation responsible for the change in expression, and that the effect of the insertion is more generally negative, in line with recent analyses done in Capsella [23], especially for long elements such as LTR-retrotransposons.

An analysis of the TIP and SNP-eQTL location with respect to the associated genes shows that while SNP-eQTLs are more evenly distributed across the genic region (Figure 1F), TIP-eQTLs are over-represented in the 1 kb region upstream of the gene (Figure 1E), which is the region that most frequently contains promoter elements regulating transcription. This suggests that although some TIP-eQTLs could be in linkage disequilibrium with the causal mutation for the change of expression, an important fraction of TIP-eQTLs could be the actual causal mutation inducing expression variation.

It has been proposed that an important fraction of gene expression variation is deleterious, and therefore, alleles associated with major expression effects should be maintained at lower frequencies [24,25]. An analysis of the effect size of the different TIP-eQTLs taking into account their frequency in the population shows that, indeed, low and high-frequency TIPs (corresponding to rare alleles) show the highest positive or negative effect on the expression of the associated gene (Figure 2 A-D, Figure 2-figure supplement 1). This is consistent with what is described as the rare variant effect [25] where rare (and likely deleterious) mutations in a population are associated with greater effects on gene expression in a population.

Effect size and population frequencies of wet and drought-stress TIP-eQTLs.

Representation of the effect size (beta) of the TIP-eQTLs with respect to their population frequency in indica (A, C) and japonica (B,D) for positive (A,B) and negative (C,D) effects. Venn diagram illustrating the intersection between TIP-eQTLs detected in each condition of the two subspecies analyzed, indica (ind), and japonica (jap) (E). Percentage of condition-specific (F) and species-specific (G) TIP-eQTLs. Shared TIPs are those falling in the intersection between the FDR-corrected TIP-eQTLs found in a given population/condition and all the associations of the other population/condition using a relaxed cutoff (p < 0.05, no FDR correction). Relationship between the frequencies in indica and japonica populations of the and 829 TIP-eQTLs (H) and 33,389 non-eQTL TIPs (I).

A comparison of TIP-eQTLs associated with changes of expression in the two different growth conditions (wet and drought stress) shows that there is an important overlap of eQTLs between the two environments (38.5 % for indica TIP-eQTLs and 33.2% for japonica TIP-eQTLs) (Figure 2E). Such overlap increases to 90% (average of indica and japonica, wet) and 71 % (average of indica and japonica, drought stress) when we compare TIP-eQTLs of one condition with non-FDR-corrected TIP-eQTLs in the other environment (Figure 2F). All the associations shared in the two environments had the same effect type on expression (positive or negative). This result suggests that the majority of TIP-eQTLs are general associations with changes of expression in both growth conditions, and that only a small number of TIP-eQTLs (81 in indica and 78 in japonica) are associated with stress-specific changes of expression. Nevertheless, among the TIP-eQTLs associated with changes of gene expression in both wet and drought stress, there are TIP associated with genes known to be involved in drought tolerance (Supplementary file 1). As an example, TIP_49046 is associated with a reduction of the expression of the gene synaptotagmin-5 (OsSYT-5). OsSYT-5 encodes a Ca2+ sensing protein with a C2 domain, is expressed in both stressed and non-stressed plants, and it has been recently shown that its silencing enhances drought tolerance in rice [26]. Other examples are TIP_23764 and TIP_52367, associated with an increased expression of the gene encoding the OsSAPK10 ABA-activated protein kinase and the S-type euonymus-related lectin gene OsEULS2 in respectively, two genes known to mediate drought stress in rice [27,28]

TIP-eQTLs are present at different population frequencies in indica and japonica

Although the overlap of TIPs associated with changes in expression under wet and drought stress conditions is very high, the overlap between indica and japonica TIP-eQTLs is low, and only 90 TIP-eQTLs out of the 563 from indica and 356 from japonica are significantly associated with variation of gene expression in both subspecies (Figure 2E). Even when we make the less-stringent comparison of the TIP-eQTLs of one subspecies with the non-FDR-corrected TIP-eQTLs of the other, as much as ~69% of indica and ~60% of japonica TIP-eQTLs appear as subspecies-specific (Figure 2F). Regarding the common associations between subspecies, all of them have the same effect type on expression (positive or negative). This result may suggest differences in the transcriptional networks in indica and japonica or, alternatively, that different alleles showing expression level differences may be differentially represented in the two subspecies. An analysis of the frequencies in indica and japonica of all the TIP-eQTLs described here shows that approximately one-third of the TIPs are only found in one of the two subspecies at MAF > 3%, although only 59 (indica) and 21 (japonica) associations correspond to truly species-specific TIPs (completely absent in one of the two subspecies). We looked for these TIPs in the rufipogon/nivara varieties and found that 56% of the indica-specific and 24 % of the japonica-specific TIPs were present in such population. This suggests that TEs have been actively inserting (or have been excised or eliminated from the population) very recently during rice evolution, likely post-domestication and during the crop diversification phase, as has already been proposed for some rice TE families [29,30]. Here we show that some of these recent TIPs are associated with changes in gene expression and may therefore have phenotypic consequences. Some of these insertions are found at high frequencies in its corresponding subspecies, which suggest that they may have been under positive selection (Supplementary file 3).

However, the vast majority of TIP-eQTLs found in this study (90%) correspond to TIPs present in both subspecies, which suggests that these insertions are relatively old and were already present in the ancestor of indica and japonica rice. Remarkably, in most cases they are found at very different frequencies in the two subspecies (Figure 2H). This suggests that these insertions were already present in the ancestor of indica and japonica rice and have been differentially retained in the two subspecies. Interestingly, an analysis of all non-eQTL-TIPs shows that an important fraction is found at the same frequency in both genomes (Figure 2I). This differential pattern between eQTL and non-eQTL TIPs is also found when we consider only the upstream gene regions (1Kb, Figure 2-figure supplement 2), suggesting that the preferential retention of TIP-eQTLs in indica or japonica may be a specific phenomenon potentially linked to their impact on the expression of the associated genes.

We examined whether TIP-eQTLs present in both indica and japonica varieties are also found in either Oryza rufipogon and Oryza nivara, the wild rice relatives believed to be the ancestors of domesticated rice [31]. We looked for the presence of the TIPs identified in rice in a set of 72 accessions of O. rufigogon and 10 accessions of O. nivara (collectively called rufipogon/nivara from now on) (Supplementary file 4). Up to 552 of the 829 TIP-eQTLs present in indica and/or japonica (66%), can be found in rufipogon/nivara varieties, confirming that an important fraction of TIP-eQTLs found in indica and japonica may be relatively old insertions already present in the wild ancestors of domesticated rice. Not surprisingly, in most cases the TIP frequencies in rufipogon/nivara are different from the frequencies of these insertions in indica and/or japonica (Supplementary file 5); these may have arisen either due to selection or genetic drift accompanying the bottleneck associated with crop evolution.

In order to identify possible selection of TIPs in indica and japonica populations, we used the Population Branch Statistic (PBS) method [32] which examines strong differentiation in frequencies between populations. This approach consists in comparing the three pairwise FST values between indica, japonica and rufipogon/nivara to estimate the frequency change in TIPs that occurred since the divergence of the two rice subspecies from its wild ancestor. We calculated PBS values for a total of 11,698 TIPs present in the japonica, indica and rufipogon/nivara populations, which included 354 TIPs that are eQTLs in in indica and/or japonica populations. TIP-eQTLs showed much higher PBS values in comparison to those not identified as eQTLs (Figure 3A) (mean absolute PBS of 0.06 vs 0.02, respectively, p < 0.01, Wilcoxon test). Furthermore, TIPs with the most extreme PBS values (above 95 percentile) are enriched for TIP-eQTLs (Fisher’s Exact test odds ratio = 4.87, p < 0.01), suggesting that a high fraction of TIP-eQTLs have been positively selected in indica or japonica. Note that some of the TIPs not identified as eQTLs in this particular dataset may actually be linked to variation of gene expression in organs, developmental stages or environmental conditions different from the ones analyzed here. Interestingly, up to 35% of the TIP-eQTLs identified here, associated with 119 genes, fell within the top 10% of absolute PBS values for all TIPs (Fig 3A), suggesting that TIPs associated with gene expression variation have likely been under differential selection during rice evolution.

Signatures of positive selection on TIP-eQTLs.

A, Absolute PBS of 354 TIP-eQTLs (left) and 11,344 TIPs (no-eQTL, right) present in indica, japonica and rufipogon populations. Dotted vertical lines represent the 95th an 99th percentile of the PBS values of the whole dataset (11,698 TIPs). Red dots represent two examples of TIP-eQTLs with extreme PBS values. B) Population frequency of the two TIPs with extreme PBS values, marked as red dots in the left panel A, as well as of the whole TIP dataset. C) Fst-based tree of the two TIPs with extreme PBS values, as well as of the whole TIP dataset (average Fst).

We observe that TIPs can show evidence of selection in japonica, indica or both subspecies. As examples, Figure 3B and C shows the representation of the PBS analysis, and the frequency in the populations, of two TIP-eQTL whose frequency has greatly increased in indica (TIP_53500), or japonica (TIP_72732), compared with the mean PBS for all TIPs.

Gene variants selected during rice evolution: some examples

Genes linked with TIP-eQTLs showing extreme indica or japonica PBS metrics are good candidates for genes underlying adaptation between different rice subspecies. Among the genes with TIP-eQTLs that have extreme PBS values in indica or japonica we can find several examples that are known to regulate plant architecture, plant and grain development or abiotic stress responses.

Some of the most extreme PBS values are for TIPs associated with changes of expression of genes involved in the signal transduction of hormones, including brassinosteroids, ABA, ethylene, jasmonic acid (JA) and auxin (Supplementary file 5). We found four different TIP-eQTLs with high PBS values associated with the EG2/OsJAZ1 gene (Os04g0653000), a locus encoding a JA signaling repressor [33]; the four insertions are related to MITEs of the Tourist transposon superfamily. Three of the insertions (TIP_32891, TIP_32892 and TIP_32894), located upstream (3.4 kb and 2.7 kb) and downstream (150 bp) of the gene, are associated with an increase of expression of EG2, and are physically linked to one another (mean r2 value = 0.88). In contrast, the fourth insertion (TIP_32893), which is not linked to them, is present within the first intron of the gene and is associated with expression reduction.

The recent advances in characterizing the pangenome of rice and the super-pangenome that includes its wild ancestors [34–36] has allowed us to characterize the locus in indica, japonica and rufipogon/nivara. This analysis showed that EG2 is primarily found in two different haplotypes (Figure 4A). One haplotype, Hap A, contains the three MITE insertions associated with higher expression of EG2 (Figure 4A, 4B, and Figure 4-figure supplement 1), while the second haplotype, Hap B, contains the fourth MITE insertion and is associated with reduced EG2 expression (Figure 4A, 4B and Figure 4-figure supplement 1). Hap A was identified to be at high frequency in rufipogon/nivara (48%), whereas Hap B is at lower frequency (16%). The frequency of both Hap A and Hap B is slightly increased in japonica (65% and 28%, respectively). In indica, however, there is a clear reduction in the frequency of Hap A and a concomitant strong frequency increase of Hap B, associated with reduced expression in indica (16% and 83%, respectively). Interestingly, EG2/OsJAZ1 is a repressor of spikelet development [33] and the number of differentiated spikelets per panicle tends to be lower in japonica as compared with indica [37].

Selection on TIP-eQTLs associated with EG2 expression.

A, Representation of the two main EG2 haplotypes present in rice and rufipogon/nivara populations identified in the rice super-pangenome. Conserved nucleotide regions are connected by gray marks. Structural variants longer than 50bp are shown as white spaces. TIP-eQTLs are shown as red boxes. Additional TIPs are shown as white boxes. B, Boxplot representation of the expression of the two EG2 haplotypes in the indica population. Numbers inside boxplots represent the number of accessions in each group. Analysis of nucleotide diversity (π) and diversity in haplotypes homozygosity tracts (H12) for Hap A (C, E) and Hap B (D, F) in japonica (C, D) and indica (E, F) populations. The vertical dotted black line shows the position of the TIP insertion. the EG2 gene is schematically shown in blue. The horizontal dashed black line represents the mean of 1000 random permutation pulls (for p-value see Supplementary file 6).

We looked more closely for signs of selection at this specific locus, by examining levels of nucleotide diversity (π [38]) and diversity in haplotypes homozygosity tracts (H12 [39]). Consistent with the expression (and possible spikelet phenotypic differences), we found that in japonica the haplotype associated with increased EG2 expression (Hap A) was under positive selection, whereas the haplotype associated with decreased expression of EG2 (Hap B) did not show any sign of selection; this indicates that positive selection acts on individuals carrying the haplotype for high expression of EG2 in japonica (Figure 4C-D). Specifically, we observe significantly lower π and higher H12 in Hap A, associated with increased EG2 expression, while this pattern was not significant for Hap B (supplementary file 6). However, in indica the opposite pattern appears to prevail, wherein there is a stronger evidence of selection for Hap B, but less so for Hap A (Figure 4E-F). This suggests positive selection towards higher expression of EG2 in japonica, and lower expression of EG2 in indica, which could potentially explain the higher number of spikelets per panicle observed in the latter[37].

Another good candidate for genes underlying adaptation of rice is the OsGAP gene. There are two different TIP-eQTLs with high PBS values associated with changes of expression of the OsGAP gene (Os07g0500300) in japonica, whereas they do not correlate with changes in expression in indica (Figure 5-figure supplement 1). OsGAP encodes a putative GTPase activating protein, similar to CAR proteins (C2-domain abscisic acid-related proteins), that play an important role in ABA signal transduction in Arabidopsis [40]. The first TIP (TIP_50057) is located ~4 kb upstream of the OsGAP gene and is associated with an increase of expression of the gene in japonica whereas the second insertion (TIP_50059) is located within the first intron and is associated with a decrease in expression in japonica (Figure 5A-B). The analysis of the locus using the super-pangenome data shows that OsGAP is found in two different haplotypes, one containing TIP_50057 (Hap A) and the other containing TIP_50059, together with some additional structural differences (Hap B; Figure 5A). The two TIPs defining the two haplotypes are present in the rufipogon/nivara population at complementary frequencies (73% and 19%, respectively). Interestingly, the proportion of the two haplotypes in cultivated rice seem to be very different, with an increased frequency of the haplotype associated with a reduced expression of OsGAP (Hap B) in both japonica (~50%) and indica (~95%) were it reaches near fixation.

Selection on TIP-eQTLs associated with OsGAP expression.

A, Representation of the two main OsGAP haplotypes present in rice and rufipogon/nivara populations identified in the rice super-pangenome. Conserved nucleotide regions are connected by gray marks. Structural variants longer than 50bp are shown as white spaces. TIP-eQTLs are shown as red boxes. Additional TIPs are shown as white boxes. B, Boxplot representation of the expression of the two OsGAP haplotypes in the japonica population. Numbers inside boxplots represent the number of accessions in each group. Analysis of nucleotide diversity (π) and diversity in haplotypes homozygosity tracts (H12) for Hap A (C) and Hap B (D) in japonica population. The vertical dotted black line shows the position of the TIP insertion. The OsGAP gene is schematically shown in blue. The horizontal dashed black line represents the mean of 1000 random permutation pulls (for p-value see Supplementary file 6).

It has been proposed that OsGAP is a negative regulator of ABA signaling in seed germination and dormancy, and that reduced expression of OsGAP may prevent rice pre-harvest sprouting (PHS) [40]. Reduced seed dormancy is a common target for selection of cultivated varieties, but this trait may come at a cost of a higher PHS risk, which may be a problem, especially in regions where heavy rain is common during the harvest season. Therefore, the appropriate degree of dormancy may depend on the agroecological conditions in which a particular variety is grown. We see here that two independent TIPs are associated with changes of OsGAP expression in japonica but not in indica (Figure 5A, Figure 5-figure supplement 1). Interestingly, there are clear and strong signs of selection for increased expression of OsGAP in japonica, wherein Hap A was under positive selection, and Hap B was selectively neutral, indicating positive selection acts on only those individuals carrying haplotype for high expression of OsGAP in japonica (Figure 5C,D; Supplementary file 6). In contrast, there is no evidence for selection in indica varieties despite the near fixation of one haplotype. This suggests that an increased level of OsGAP may be relevant for the control of dormancy in japonica but not in indica (despite its near-fixation in the latter), which would be in line with recent data suggesting that seed dormancy is regulated by different genes/alleles in indica and japonica [41].

Finally, we find a TIP-eQTL whose frequency has greatly increased in cultivated rice with respect to rufipogon/nivara. TIP_45706, corresponding to an insertion of a gypsy-like LTR-RT at ~1 kb upstream of the OsMPH1 gene (Figure 6A). This insertion is associated with a decrease in the expression of OsMPH1 in indica and possibly japonica varieties (Figure 6B, Figure 6-figure supplement 1), and is found at low frequency in rufipogon/nivara (~11%) but at high frequency in both indica (70%) and japonica (95%). OsMPH1 encodes a MYB-like transcription factor that has been shown to regulate plant height, its reduced expression resulting in shorter plants [42]. A reduction in plant height due to a mutation in the SD1 gene, which encodes the gibberellin biosynthesis gene GA-20ox, was at the origin of the Green Revolution in the 1960s, but it has been shown that alleles of SD1 resulting in shorter culm length were also selected during the domestication of japonica rice [43]. Our results suggest possible parallel selection of alleles in other genes that may lead to similar phenotypes. This is further reinforced by strong selection acting on the haplotype containing the TIP_45706 insertion in both indica and japonica (Figure 6C and 6D Supplementary file 6), leading to reduced expression of OsMPH1 and thus shorter plants.

Selection on TIP-eQTLs associated with OsMPH1 expression.

Representation of the two main OsMPH1 haplotypes present in rice and rufipogon/nivara populations identified in the rice super-pangenome. Conserved nucleotide regions are connected by gray marks. Structural variants longer than 50bp are shown as white spaces. TIP-eQTLs are shown as red boxes. B, Boxplot representation of the expression of the two OsMPH1 haplotypes in the japonica population. Numbers inside boxplots represent the number of accessions in each group. Analysis of nucleotide diversity (π) and diversity in haplotypes homozygosity tracts (H12) for Hap A in indica population (C) and in japonica population (D). The vertical dotted black line shows the position of the TIP insertion. The OsMPH1 gene, which is schematically shown in blue. The horizontal dashed black line represents the mean of 1000 random permutation pulls (for p-value see Supplementary file 6).

Discussion

Transposons are a major source of genome variability, and are known to affect gene expression in numerous ways. The importance of rare variants and unfixed TE insertions for gene expression variation in humans [44] and plants [23] has already been recently described. A recent pangenome analysis in tomato has shown that the phenotypic variation of important crop traits is linked to structural variants present in the species, which are associated to subtle changes of transcription of genes involved in signal transduction. Many of these SV-eQTLs are likely TIPs, and some lead to phenotypic consequences such as the jointless trait caused by a transposon insertion, which allows complete separation of fruits from other floral parts [45]. In the present study we evaluated the importance of TIP-related expression variability in the recent evolution of rice. To this end we performed a TIP-eQTL mapping using expression data from rice varieties from the O. sativa ssp. indica and O. sativa ssp. japonica subspecies. We show here that using TIPs in addition to SNPs as genetic information allows the uncovering of additional genomic associations to changes in gene expression, and that when both TIPs and SNPs are both associated, TIPs often explain more of the variance in expression. This is in line with recent reports that have also used TIPs for GWAS with different phenotypes in tomato and rice [7,8], and suggests that, even if some of the TIPs described here are probably just linked to the causal mutation of the phenotype (ie, a SNP or a different type of SV), they may be more frequently the causal mutation themselves as compared with SNPs. The concentration of TIP-eQTLs in upstream regions of genes, where most transcriptional regulatory elements usually reside, also suggest that a high fraction of the TIP-eQTLs here described are the actual mutation underlying gene expression variation.

The close association of some TEs with genes could also partially explain the frequent association of more than one TIP with changes of expression of particular genes. Indeed, among the 718 genes with TIP-eQTLs described here, 18% have two or more TIPs associated with changes of expression. Interestingly, among the 30 genes with TIPs in indica and japonica that have the highest PBS – an indicator of strong differential selection - 43% have more than one TIP-eQTL. This suggests that TIPs are an important source of gene expression variability in rice, in particular in genes that may have been strongly selected during its recent evolution. In some cases, the different TIPs linked to a gene are associated with opposite effects on its expression and are present in different haplotypes. In these instances, as shown for the OsJAZ1 and OsGAP genes, the two haplotypes have strong signs of selection, although one positively and the other negatively, which suggests that both TE insertions may have played a role in the diversification of rice subspecies.

The results presented here show that most TIPs associated with variation of gene expression in indica and/or japonica are relatively old insertions that are also present in rufipogon/nivara, the wild ancestors of rice. This is in line with recent data showing that a high number of the structural variants associated with changes of expression in tomato were already present in its wild ancestor [45] and highlights the importance of the standing variation, already present in the wild ancestors, for crop adaptation. It is assumed that most TE insertions are selectively neutral or slightly deleterious [46], and the presence of TIPs associated with expression variation of genes in the wild ancestors of rice, suggests that this variation can be well tolerated in wild rice. Interestingly, we show here that many of these TIPs have been positive or negative selected in rice populations, which shows that they translate into selectable phenotypic differences in the agroecological conditions of cultivated rice. Indeed, we show examples of selected variants with modified expression of genes known to be linked to important traits that were targets of selection. It has been already proposed that a major fraction of the mutations linked to crop domestication and breeding are associated with changes of gene expression involved in signal transduction [10,11]. Here we show that specific expression variants of genes involved in signal transduction have been differentially selected in indica and japonica rice populations. In addition, our results also point to TEs as a major driver of gene expression variation selected during crop adaptation and breeding.

Methods

TIP detection

Re-sequencing data for 126 indica and 82 japonica rice accessions was obtained from Groen et al [19] (Bioproject accessions PRJNA557122, PRJNA422249 and PRJEB6180). BBDuK (https://sourceforge.net/projects/bbmap/) was used for adaptor and quality trimming. Clean reads were aligned to the Nipponbare reference genome [47] using BWA [48]. PoPoolationTE2 [20] was used to detect TIPs in the mode "joint" using Nipponbare TE annotation described by Ou et al [47]. TIPs with a zygosity lower than 0.25 in all samples were excluded to avoid false positives. TIPs were further filtered using the parameters –-min-count 5, --max-otherte-count 2, --max-structvar-count 2, and only those having MAF higher than 3% and no missing data were kept. Finally, the TIP matrix was transformed to binary form using zygosity cutoff of 0.05 to define an insertion as present.

Evaluation of TIP-calling performance in our dataset

Re-sequencing data from 48 indica rice accessions randomly sampled from our dataset were used to identify TIPs with the objective of evaluating the performance of TIP calling. MH63 was included among these accessions. This variety has a chromosome-level assembly as well as short-read re-sequencing data publicly available. MH63 sequencing short-reads were obtained from NCBI SRA accession SRX1639978 and subsampled to 15X to match the mean coverage of the full dataset. TIP detection was carried out using Nipponbare genome as reference, and the “joint” mode of PopoolationTE2, with the parameters described in the methods section. We used RepeatMasker to annotate TEs in these MH63 assembled regions, and used the annotated TEs to benchmark precision and sensitivity of the TIP calls, according to the following formulas: Sensitivity=TP/(TP+FN)Precision=TP/(TP+FP)

True positives (TP) were TIPs detected by PopoolationTE2 that could be detected in the MH63 orthologous region by RepeatMasker. False positives were TIPs detected by PopoolationTE2 that had no TE of the same family annotated by RepeatMasker in the corresponding region of MH63. False negatives (FN) were cases when the TIP prediction of MH63 was an absence, but the assembled region contained the TE detected by RepeatMasker.

TIP and SNP-eQTL mapping

Transcriptome data for the 208 rice accessions (wet and drought stress conditions; tree independent replicates per condition) was obtained from Groen el al [19] and separated into individual replicate matrices. Transcripts expressed in more than 10% of the samples on each replicate were extracted and counts were normalized using the vst function of DESeq2 [49]. TIP and SNP-eQTL mapping were performed using Matrix eQTL software [50], applying the simple linear regression model and including subpopulation groups as covariates. We used a cut-off distance of 5,000bp to identify cis-eQTLs and a 5% False Discovery Rate threshold for multiple testing corrections.

Selection analyses

For the Population Branch Statistics (PBS) analysis we looked for the presence of rice TIPs in the 82 accessions belonging to the rufipogon/nivara population (Supplementary file 4) and retained only those present in the three populations (indica, japonica, rufipogon/nivara), resulting in a matrix of 13,622 TIPs and 288 accessions. The TIP matrix was filtered to remove TIPs with more than 5% missing data, or with MAF < 5%. The remaining missing data (1.7%) was imputed using the “wright” algorithm implemented in SNPready. The clean matrix (11,698) was used to calculate the individual PBS values per TIP, following the formulas described in Yi et al., (2010)[32].

Along with PBS, to evaluate whether TIP-eQTLs show signs of selection, we estimated nucleotide diversity (pi [38]), which is a SFS (site-frequency spectrum) measure to test for the presence of selection. This was done using a custom script [51] in a window of 25SNPs with a sliding window of 5. Further, we also estimated the H12 homozygosity statistic [39] that can detect the presence of both hard and soft sweeps associated with selection. Since we expect long runs of homozygosity in the regions around the selected loci, H12 was calculated in windows of 50 SNPs, with a sliding window of 25. Both these statistics were estimated separately for indica and japonica, and for the samples with TIP insertions present and absent. This was for a range of 100kb (50kb up-and downstream the TIP insertion), using SNPs identified from Groen et al [19] but without the 1000bp linkage thinning. To test for the significance of the statistics, we performed a permutation test (N=1000) with random sampling with the same number of individuals that were used to estimate the selection statistics. The statistics were estimated for these 1000 random pulls in the same regions, and using the same window sizes as the original statistics.